---
title: "Health and economic effects of weather events"
author: "FD"
date: "29 April 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Health and economic effects of weather events

## Summary

## Data processing

### Data loading

The data is downloaded from the course website if not existing in the current working directory.

Information about the data can be found at the following locations:

- [National Weather Service Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
- [National Climatic Data Center Storm Events FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

According to the documentation, just direct fatalities and injuries are part of the database. Indirect fatalities and injuries are not reported in its own field.

```{r}
if(!file.exists("repdata-data-StormData.csv.bz2")) {
  download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2")
}
#data=read.csv(bzfile("repdata-data-StormData.csv.bz2"))
```

### Data clean up

Data column names are normalized to lower case letters.

```{r}
library(dplyr)

colnames(data)=tolower(colnames(data))
```

Many event types in the data set are doubled like: Damaging Freeze, DAMAGING FREEZE but also more complex cases like: DOWNBURST, DOWNBURST WINDS. To encount for the simplest cases, the event types are normalized to lower case before the analysis is done.

```{r}
data$evtype<-tolower(data$evtype)
```
  
Additionally, two data sets are prepared which fits the need of the following analysis.

```{r}
eventData=select(data, evtype, fatalities, injuries)
```

## Most harmful events with respect to population health

Population health is summarized by fatalities and injuries. For each of these categories, there is a column in the data set.

To answer the question which events are most harmful, the data has to be summarized by event type.

```{r}
library(reshape2)

molten=melt(eventData, id.vars=c("evtype"))
averagedByEventtype=dcast(molten, evtype ~ variable, sum)
```

Now, lets sort the events and look at the cummulative distribution. Lets concentrate in the following on the events which causes 90% of the fatalities or injuries.

```{r}
library(ggplot2)
library(plotly)

orderedByFatalities=averagedByEventtype[sort(averagedByEventtype$fatalities, index.return=T, dec=T)$ix, ]
orderedByInjuries=averagedByEventtype[sort(averagedByEventtype$injuries, index.return=T, dec=T)$ix, ]

incrementedFatalities=Reduce(function(x, y){append(x, y+tail(x, n=1))}, orderedByFatalities$fatalities, 0)

topFatalities=orderedByFatalities[summed<0.9*tail(summed, n=1),]

qplot(ordered(topFatalities$evtype, levels=topFatalities$evtype), topFatalities$fatalities)+geom_bar(stat = "identity")

topInjuries=orderedByInjuries[orderedByInjuries$injuries>quantile(orderedByInjuries$injuries, 0.9),]
```

## Events with the greatest economic consequences

## Results
